%!TEX root = ../LLMSeminar.tex
% ──────────────────────────────────────────────────────────────
%  Section 2 — The Illusion of Chat
% ──────────────────────────────────────────────────────────────

\section{The illusion of chat}\label{sec:chat-illusion}

\subsection{A single API call}

You interact with an LLM via a single API call.  One string in, one
string out---that is the entire interaction.

\begin{center}
\begin{tikzpicture}
  \node[sysbox fill, minimum width=3.5cm] (client) {\textbf{Client}};
  \node[sysbox fill, minimum width=3.5cm, right=6cm of client] (api) {\textbf{API}};

  \draw[sysarrow] ([yshift=0.4cm]client.east) -- ([yshift=0.4cm]api.west)
    node[midway, above, font=\small] {prompt (string)};
  \draw[sysarrow] ([yshift=-0.4cm]api.west) -- ([yshift=-0.4cm]client.east)
    node[midway, below, font=\small] {completion (string)};
\end{tikzpicture}
\end{center}

At the protocol level, this is an HTTP POST request carrying a JSON
payload:

\begin{lstlisting}
POST /v1/messages
{
  "model": "claude-sonnet-4-6",
  "messages": [
    {"role": "user",
     "content": "What is a quark?"}
  ]
}
\end{lstlisting}

That is it.  There is no session, no connection state, no memory.

\subsection{How chat works}

The multi-turn ``conversation'' you experience in ChatGPT or Claude is
an illusion maintained entirely by the \emph{client}.  The client
re-sends the \textbf{entire} conversation history with every turn:

\begin{lstlisting}
Turn 1:  [user_1]
Turn 2:  [user_1, asst_1, user_2]
Turn 3:  [user_1, asst_1, user_2, asst_2, user_3]
\end{lstlisting}

Each turn is an independent function call.  The model receives the
accumulated transcript as input and produces its next response as
output.  It has no server-side memory of previous turns.

\begin{keyresult}[Chat is a client-side loop]
  The LLM has no memory.  The client maintains the illusion.  Chat is a
  loop; memory is a fiction.
\end{keyresult}

\subsection{The context window}

Every model has a finite \textbf{context window}---the maximum number of
tokens it can process in a single call.

\begin{center}
\begin{tikzpicture}
  \fill[isabelline] (0,0) rectangle (11,1);
  \fill[cgblue!20] (0,0) rectangle (1.8,1);
  \node[font=\scriptsize, text=spacecadet, align=center] at (0.9,0.5)
    {system\\prompt};
  \fill[munsell!15] (1.8,0) rectangle (8.5,1);
  \node[font=\scriptsize, align=center] at (5.15,0.5)
    {conversation history};
  \fill[munsell!30] (8.5,0) rectangle (11,1);
  \node[font=\scriptsize, align=center] at (9.75,0.5)
    {latest\\message};
  \draw[spacecadet, line width=0.8pt] (0,0) rectangle (11,1);
  \draw[decorate, decoration={brace, amplitude=6pt, mirror},
        spacecadet, line width=0.8pt]
    (0,-0.2) -- (11,-0.2)
    node[midway, below=8pt, font=\small] {context window (tokens)};
\end{tikzpicture}
\end{center}

Typical sizes: $8\text{k}$ tokens for free chat, $128\text{k}$ to
$200\text{k}$ for frontier models, and up to $1\text{M}$ tokens for the
largest available context windows.  When the conversation exceeds the
window, something must be dropped.

\begin{remark}
  A full PhD thesis fits comfortably in a $1\text{M}$-token context
  window.  A thesis chapter fits in roughly $20\%$ of a
  $200\text{k}$-token window.
\end{remark}

\subsection{Context rot}

As the context fills, older information gets buried.  The model's
effective attention degrades.  Your constraints, instructions, and
earlier results silently disappear.

\begin{keyresult}[Why chat fails]
  This is why chat interfaces silently ruin your results.  Context rot
  is the primary failure mode of naive LLM usage.
\end{keyresult}

\subsection{The system prompt}

A hidden preamble is prepended to every call.  It shapes behaviour,
persona, and constraints.  But it is \emph{just more text}---there is
nothing magical about it.  Anthropic's system prompt has been published;
it is plain English\footnote{%
  \url{https://docs.anthropic.com/en/release-notes/system-prompts}}.

\subsection{Summary}

Chat is a loop.  Memory is an illusion maintained by the client.  The
context window is finite and when it fills, your results degrade.  Every
interaction with an LLM is, at bottom, one string in, one string out.
