%!TEX root = ../LLMSeminar.tex
% ──────────────────────────────────────────────────────────────
%  Section 1 — What is an LLM?
% ──────────────────────────────────────────────────────────────

\section{What is an LLM?}\label{sec:what-is-llm}

\subsection{From outside, looking in}

Let us set aside everything you have heard about neural networks,
training data, and alignment.  From outside the API boundary, the
observable behaviour of a large language model is captured by a single
type signature:

\begin{definition}[LLM as a function]\label{def:llm-function}
  A large language model is a function
  \[
    f : \Str \to \Str.
  \]
  It accepts a string and returns a string.  No hidden state is carried
  between calls.
\end{definition}

That is the complete external specification.  Everything else---the
transformer architecture, the training procedure, the reinforcement
learning from human feedback---is implementation detail.  Important
implementation detail, but detail nonetheless.

\subsection{Stateless}

Each call to the LLM is independent.  There is no server-side memory
that persists between invocations.  Call~1 and Call~2 share nothing:

\begin{center}
\begin{tikzpicture}
  \node[sysbox fill] (c1) {Call 1};
  \draw[sysarrow] ([xshift=-2cm]c1.west) -- (c1.west)
    node[syslabel] {input};
  \draw[sysarrow] (c1.east) -- ([xshift=2cm]c1.east)
    node[syslabel] {output};

  \node[sysbox fill, right=4cm of c1] (c2) {Call 2};
  \draw[sysarrow] ([xshift=-2cm]c2.west) -- (c2.west)
    node[syslabel] {input};
  \draw[sysarrow] (c2.east) -- ([xshift=2cm]c2.east)
    node[syslabel] {output};

  \draw[dashed, color=spacecadet!25, line width=1pt]
    ([xshift=1.5cm]c1.east) -- ([xshift=-1.5cm]c2.west)
    node[midway, font=\huge, text=red!70!black] {$\times$};
\end{tikzpicture}
\end{center}

This is a crucial point.  When you close a chat window and open a new
one, the model has genuinely forgotten everything.  It is not pretending.
It is stateless.

\subsection{Nondeterministic}

The same input can produce different outputs.  This is not a bug---it is
by design.  The model \emph{samples} from a probability distribution
over possible next tokens:
\[
  f \;\text{samples from}\; \Prob(\text{next token} \given \text{context}).
\]

Feed the prompt \code{"What is spin?"} three times and you may receive
three different explanations.  The model is a nondeterministic function.

\subsection{Temperature}\label{subsec:temperature}

The probability that the model selects token~$i$ is given by the
softmax with a temperature parameter~$T$:

\begin{keyresult}[Token sampling distribution]
  \[
    \Prob(\text{token}_i) \;=\; \frac{\exp(\ell_i / T)}
      {\displaystyle\sum_j \exp(\ell_j / T)},
  \]
  where $\ell_i$ is the logit (raw score) for token~$i$.
\end{keyresult}

\begin{intuition}[Boltzmann analogy]
  This is the Boltzmann distribution.  It is exactly the formula for the
  probability of a microstate with energy~$-\ell_i$ at temperature~$T$
  in statistical mechanics.  The two limits are:
  \begin{itemize}
    \item $T \to 0$: deterministic---the model always picks the
      highest-scoring token (ground state).
    \item $T \to \infty$: uniform---every token is equally likely
      (infinite temperature).
  \end{itemize}
  The temperature knob is a dial between ``play it safe'' and
  ``be creative.''
\end{intuition}

\subsection{Tokens}

The atomic unit is not a character and not a word, but a \emph{token}:
a subword chunk drawn from a fixed vocabulary~$V$ of approximately
$100{,}000$ entries.

\begin{example}[Tokenisation]
  \begin{lstlisting}
'theoretical physics'  ->  ['theor', 'etical', ' physics']
  \end{lstlisting}
\end{example}

More precisely, the function signature is
$f : \Tok^* \to \Tok^*$,
where $\Tok^*$ denotes a finite sequence of tokens.

\subsection{Inference}

An LLM call is a single forward pass through a frozen neural
network.  There is no learning and no state mutation during
inference.  The weights are fixed at deployment time; the model does not
update itself when you talk to it.

The neural network inside is a \textbf{Transformer}
\cite{vaswani2017attention}.

\begin{greybox}[Further reading]
  \begin{itemize}
    \item Vaswani et al., \emph{Attention Is All You Need} (2017).
    \item Andrej Karpathy, \emph{nanoGPT}---build a GPT from scratch.
    \item Jay Alammar, \emph{The Illustrated Transformer}.
    \item PicoGPT.jl---a minimal GPT implementation in Julia.
  \end{itemize}
\end{greybox}

\subsection{Summary}

\begin{keyresult}[Definition of an LLM]
  An LLM is a \textbf{stateless, nondeterministic function}
  \[
    f : \Str \to \Str
  \]
  that samples from $\Prob(\text{next\_token} \given \text{context})$
  via a single forward pass through a frozen neural network.
\end{keyresult}
